\providecommand{\main}{..}
\documentclass[\main/notes.tex]{subfiles}

\begin{document}
	\chapter{Graphics Systems and Models}
		\begin{definition}{Computer Graphics}
			All aspects of producing pictures or images using a computer.

			Began 50 years ago, with the display of lines on a \concept{cathode-ray tube (CRT)}.
		\end{definition}
		\begin{definition}{WebGL}
			A graphics software system supported by most modern web browsers.
			A version of \concept{OpenGL} -- the widely accepted standard for developing graphics applications.
		\end{definition}
		\section{Applications of Computer Graphics}
			\begin{sidenote}{Four Major Areas}
				\begin{center}
					\begin{enumerate*}[itemjoin=\quad]
						\item Display of information
						\item Design
						\item Simulation and animation
						\item User interface
					\end{enumerate*}
				\end{center}
			\end{sidenote}
			\subsection{Display of Information}
				\begin{example}[Types]
					\begin{itemize}[nosep]
						\item Floor plans
						\item Maps
						\item Statistics Plots
						\item Medical Imaging\\
							\begin{itemize*}[itemjoin=\quad]
								\item Computed Tomography (CT)
								\item Magnetic Resonance Imaging (MRI)
								\item Ultrasound
								\item Positron-emission tomography (PET)
							\end{itemize*}
						\item Scientific visualisation
					\end{itemize}
				\end{example}
			\pagebreak
			\subsection{Design}
				Starting with a set of specifications, enginees and architects seek a cost-effective and aesthetic solution that satsifies the specifications. Design is an iterative process.

				Design problems are either \concept{overdetermined} or \concept{underdetermined}.
				\begin{description}[nosep]
					\item[Overdetermined] No solution that satisfies all the criteria.
					\item[Underdetermined] Multiple solutions that satisfy all the criteria.
				\end{description}

				Using interactive graphic tools in \concept{computer-aided design (CAD)} occurs in many fields.
			\subsection{Simulation and Animation}
				An example is the training of pilots -- using graphical flight simulators increases safety, and reduces training expenses.

				The success of simulators led to using computer graphics for animation, as well as VR.
			\subsection{User Interfaces}
				The predominant way of interacting with computers uses a visual paradigm that includes windows, icons, menus, and a pointing device.
\end{document}